A Heuristic Approach towards Drawings of Graphs with High Crossing Resolution

### 2018-08-15 paper/comments/comments.tex

\tcomment{However the approach is not novel (although it has been presented as such) - and so the paper comes across as preliminary work that needs to be rooted in a wider context in order to strengthen it.}

\tcomment{The process of choosing whether to move a node ("If there exists a position among them, that does not lead to a reduction of the crossing resolution, we move the vertex to this position") is an application of a simple hill-climbing algorithm (with the same advantages and disadvantages) - this needs to be acknowledged and clarified.}

@@

\tcomment{Five aesthetic properties (P1-P5) are clearly defined, but never referred to again. There are only four comparative line charts in each of fig 3 and fig 4 (one of which - iterations - is not an aesthetic property).  P2 and P3 do not seem to have been measured or plotted.}

@@

\tcomment{l.103: The result of Dujmovic et al. looks like folklore. Split the edge set of the drawn graph into $\pi/\alpha$ parts each of which must be planar. If this is indeed correct, you can perhaps mention it.}

@@

\tcomment{l.141: "... weighted random selection procedure ..." Could you be more specific?}

@@

\tcomment{l.193: How big is a medium-size graph?}

@@

\tcomment{Improve the explanation at the end of the section "Complexity issues" on page 6.

@@

\tcomment{Categorization of aesthetic properties for evaluating the present algorithm

@@

>>>>

\rcomment{However the approach is not novel (although it has been presented as such) - and so the paper comes across as preliminary work that needs to be rooted in a wider context in order to strengthen it.}

\response{See below.}

\rcomment{The process of choosing whether to move a node ("If there exists a position among them, that does not lead to a reduction of the crossing resolution, we move the vertex to this position") is an application of a simple hill-climbing algorithm (with the same advantages and disadvantages) - this needs to be acknowledged and clarified.}

\response{Added a note stating that our approach essentially resembles a hill climbing method.}

@@

\rcomment{Five aesthetic properties (P1-P5) are clearly defined, but never referred to again. There are only four comparative line charts in each of fig 3 and fig 4 (one of which - iterations - is not an aesthetic property).  P2 and P3 do not seem to have been measured or plotted.}

\response{Removed the notation with P.}

@@

\rcomment{l.103: The result of Dujmovic et al. looks like folklore. Split the edge set of the drawn graph into $\pi/\alpha$ parts each of which must be planar. If this is indeed correct, you can perhaps mention it.}

@@

\rcomment{l.141: "... weighted random selection procedure ..." Could you be more specific?}

\response{Changed to probabilistic.}

@@

\rcomment{l.193: How big is a medium-size graph?}

\response{Made a note that we consider large to be more than 100 vertices.}

@@

\rcomment{Improve the explanation at the end of the section "Complexity issues" on page 6.

@@

\response{Refined the text a bit. No new figure.}

\rcomment{Categorization of aesthetic properties for evaluating the present algorithm

@@

\response{Pointed out that P4 and P5 are additionally measured.}

### 2018-08-15 paper/paper.tex

Formally, the value of the minimum angle formed by any two crossing edges in a drawing is referred to as its \emph{crossing resolution}; the crossing resolution of a graph is defined as the maximum crossing resolution over all its drawings. Clearly, the crossing resolution of a non-planar graph is at most $90^\circ$, while a graph that admits a drawing with crossing resolution $90^\circ$ is called \emph{right-angle-crossing} graph or \emph{RAC} graph, for short; see Figure~\ref{fig:examples}. For these graphs, several results, mostly of theoretical nature, are known (refer to Section~\ref{sec:relatedwork} for a short overview). Notably, RAC graphs are sparse (they contain at most $4n-10$ edges~\cite{DBLP:journals/tcs/DidimoEL11}, where $n$ denotes the number of  vertices), while deciding whether a graph is RAC is NP-hard~\cite{DBLP:journals/jgaa/ArgyriouBS12}.

@@

The role of critical vertices is central in our algorithm
%\footnote{If the focus is not on the critical vertices for a large graph, then an algorithm that is based on randomly selecting a vertex to move, so to improve the crossing resolution, will need a large number of iterations to converge to a good solution, because it is very unlike to select one of them to move.}
\footnote{If the focus is not on the critical vertices for a large graph, then our algorithm will need a large number of iterations to converge to a good solution, because it is simply very unlikely to select to move one of the vertices that define the crossing resolution.}: By appropriately changing the location of a critical vertex or of a vertex in the neighbourhood of the critical vertices, we naturally expect to improve the crossing resolution of the current drawing. We turned this observation into an algorithmic implementation through a weighted random selection procedure, so that the vertices at distance~$i$ from the ones of the vertex-pool (that is the shortest path to any vertex in the vertex-pool has length~$i$)  have higher weights than the corresponding ones at distance $j$  in the graph, when $0 \leq i<j$. So, if the vertex-pool contains only critical vertices, then the closer a vertex is to the critical vertices, the more likely it is to be chosen. Otherwise, the vertex-pool contains all vertices and each vertex can be chosen with the same probability.

@@

So far, we have described the main idea of our algorithm, which at each iteration chooses uniformly at random a vertex of the current drawing to move (based on the content of the vertex-pool), so to improve the crossing resolution. Next, we  describe how to compute its new position in the next drawing.

@@

If none of the points $\pi_j$, with $j=0,1,\ldots,\rho-1$ is feasible, then the position of $v_i$ in $\Gamma_i$ is $p(v_i)$, i.e., same as in $\Gamma_{i-1}$, since $c(\Gamma_i) \geq c(\Gamma_{i-1})$ must hold. If there is one or more feasible points, then one may consider two different approaches to determine the position of $v_i$ in $\Gamma_i$. The most natural is to choose the feasible point that maximizes the crossing resolution of the obtained drawing. As an alternative, one may rely again on randomization and chose uniformly at random one of the feasible points as the position of $v_i$ in $\Gamma_i$. We note that we did not observe any significant difference between these two approaches (in terms of the crossing resolution of the obtained drawings), so we simply adopted the first one.

The termination condition of our algorithm is simple and depends on an in\-put parameter~$\tau$. More specifically, if the crossing resolution has not improved~during the last $\tau$ iterations, we assume that the algorithm has converged and we~stop.

% ==================================================================
\myparagraph{Avoiding local maxima}
% ==================================================================
%
To avoid getting trapped to locally optimal solutions, we mainly investigated two approaches, which are both  parametrizable by two input parameters $\zeta$ and $\zeta'$. The first mimics the human behaviour. What would~one do to escape from a locally optimal solution? She would stop trying to move the endvertices of the edges defining the crossing resolution; she would rather start moving ``irrelevant'' vertices hoping that
%by changing the embedding of the graph or its crossing structure,
by doing so a better solution will be easier to be computed afterwards. Our algorithm is mimicking this idea as follows:
%
\begin{inparaenum}[(i)]
%\item As long as the algorithm detects that within the last $\zeta$ iterations the crossing resolution of the graph has being improved, the vertex-pool contains only critical vertices.
\item if during the last $\zeta$ iterations the crossing resolution has not been~improved, then the vertex-pool becomes \emph{wider} by including all the vertices, and the algorithm is executed with this vertex-pool for $\zeta'$ iterations;
\item afterwards, the vertex-pool switches back to the critical vertices.
\end{inparaenum}
%
%From our extensive practical analysis, we quickly realized that, while
While this approach turned out to be quite effective for medium-size graphs, for larger graphs, unfortunately, it was not so efficient; in most iterations with the wider vertex-pool, the embedding could not change in a beneficial way for the algorithm~to~proceed.

@@

If the algorithm had to compute all crossing points and the corresponding angles for each candidate position of each iteration, then it would not be useful. Instead, we adopted a different approach, which turned out to be quite efficient in practice. Recall that we denoted by $v_i$ the vertex chosen at the $i$-th iteration step, and by $\pi_0,\ldots,\pi_{\rho-1}$ the candidate points to move $v_i$. Let $e_0,\ldots,e_{d_i-1}$ be the edges incident to $v_i$, where $d_i=deg(v_i)$. Next, for each edge $e_k$ with $k=0,\ldots,d_i-1$  we compute the crossings and the corresponding crossing angles of $e_k$ with all other edges in $\Gamma_{i-1}$. Let $\phi_i$ be the minimum crossing angle computed; this is our reference angle. Also, for each candidate position $\pi_j$ with $j=0,\ldots,\rho-1$, and for each edge $e_k$ with $k=0,\ldots,d_i-1$, we compute the crossings and the corresponding crossing angles of $e_k$ with all other edges of the drawing, assuming that $v_i$ is at $\pi_j$. Let $\chi_j$ be the minimum crossing angle computed with this approach, when $v_i$ is at position $\pi_j$. Clearly, $\pi_j$ is feasible only if $\chi_j \geq \phi_i$. Note that the complexity of this approach is $O(deg(v_i)m) = O(nm)$.

@@

In this section, we present the results of our experimental evaluation. For comparison purposes, apart from our algorithm, we also implemented the force-directed algorithms of Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13} and Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}. The implementations\footnote{Our implementation is publicly available at \url{https://github.com/michaelbekos/crossingresolution}} were in Java using yFiles~\cite{DBLP:books/sp/04/WieseE004}. The experiment was performed on a Linux laptop with four cores at 2.4 GHz and 8 GB RAM.
%
As a test set for our experiment, we used the non-planar Rome graphs~\cite{DBLP:reference/crc/BattistaD13}, which form a collection of around 8.100 benchmark graphs; in Appendix~\ref{app:north}, we also report on the AT\&T graphs.

The experiment was performed as follows. Initially, each Rome graph was laid out using the SmartOrganic layouter of yFiles~\cite{DBLP:books/sp/04/WieseE004}. Starting from this layout, every graph was drawn with
\begin{inparaenum}[(i)]
\item our algorithm,
\item our algorithm restricted~not to violate the aspect ratio of the initial layout, and the force-directed algorithms
\item by Argyriou et al.\ and
\item by Huang et al.
\end{inparaenum}
%
We compared the quality of the produced drawings based on the following aesthetic~properties:
%
\begin{multicols}{2}
\begin{enumerate}[P.1.]
\item \label{p:cr-res} crossing resolution
\item \label{p:to-res} total resolution
\item \label{p:an-res} angular resolution
\columnbreak
\item \label{p:as-rat} aspect ratio
\item \label{p:no-xing} number of crossings
\end{enumerate}
\end{multicols}
%
Since all algorithms of the experiment can easily be adjusted to maximize~only the crossing resolution, or only the angular resolution or both (by maximizing~the total resolution), for~P.\ref*{p:cr-res}, P.\ref*{p:to-res} and P.\ref*{p:an-res}, we adjusted each of the algorithms to maximize exclusively the corresponding measures; see Figs.~\ref{fig:cr-res}, \ref{fig:to-res} and~\ref{fig:an-res}. In our algorithm, this can be achieved by modifying appropriately the content of the vertex-pool (as we saw in Section~\ref{ssec:variants}), while in the algorithms of Argyriou et al.\ and of Huang et al.\ by switching on only the forces that maximize the corresponding properties under measure (note that, each of these two algorithms has a different set of forces to maximize the crossing and the angular resolution, such that together they maximize the total resolution). The reported results are on average across different drawings with same number of vertices.

>>>>

The value of the minimum angle formed by any two crossing edges in a drawing is referred to as its \emph{crossing resolution}; the crossing resolution of a graph is defined as the maximum crossing resolution over all its drawings. Clearly, the crossing resolution of a non-planar graph is at most $90^\circ$, while a graph that admits a drawing with crossing resolution $90^\circ$ is called \emph{right-angle-crossing} (\emph{RAC}) graph; see Figure~\ref{fig:examples}. Notably, RAC graphs are sparse with at most $4n-10$ edges~\cite{DBLP:journals/tcs/DidimoEL11}, while deciding whether a graph is RAC is NP-hard~\cite{DBLP:journals/jgaa/ArgyriouBS12}.

@@

The role of critical vertices is central in our algorithmwiki\footnote{If the focus is not on the critical vertices for a large graph, then our algorithm will need a large number of iterations to converge to a good solution, because it is simply very unlikely to select to move one of the vertices that define the crossing resolution.}: By appropriately changing the location of a critical vertex or of a vertex in the neighbourhood of the critical vertices, we naturally expect to improve the crossing resolution of the current drawing. We turned this observation into an algorithmic implementation through a probabilistic random selection procedure, so that the vertices at distance~$i$ from the ones of the vertex-pool (that is the shortest path to any vertex in the vertex-pool has length~$i$)  have higher probability for selection than the corresponding ones at distance $j$  in the graph, when $0 \leq i<j$. So, if the vertex-pool contains only critical vertices, then the closer a vertex is to the critical vertices, the more likely it is to be chosen. Otherwise, the vertex-pool contains all vertices and each vertex can be chosen with the same probability.

@@

So far, we have described the main idea of our algorithm, which at each iteration chooses uniformly at random a vertex of the current drawing to move (based on the content of the vertex-pool), so to improve the crossing resolution. Note that this method resembles probabilistic hill climbing approaches. Next, we  describe how to compute its new position in the next drawing.

@@

If none of the points $\pi_j$, with $j=0,1,\ldots,\rho-1$ is feasible, then the position of $v_i$ in $\Gamma_i$ is $p(v_i)$, i.e., same as in $\Gamma_{i-1}$, since $c(\Gamma_i) \geq c(\Gamma_{i-1})$ must hold. If there is one or more feasible points, then one may consider two different approaches to determine the position of $v_i$ in $\Gamma_i$. The most natural is to choose the feasible point that maximizes the crossing resolution of the obtained drawing. As an alternative, one may rely again on randomization and chose uniformly at random one of the feasible points as the position of $v_i$ in $\Gamma_i$. We note that we did not observe any significant difference between these two approaches (in terms of the crossing resolution of the obtained drawings), so we simply adopted the first one. 
The termination condition of our algorithm is simple and depends on an in\-put parameter~$\tau$. More specifically, if the crossing resolution has not improved~during the last $\tau$ iterations, we assume that the algorithm has converged and we~stop.

% ==================================================================
\myparagraph{Avoiding local maxima}
% ==================================================================
%
To avoid getting trapped to locally optimal solutions, we mainly investigated two approaches, which are both  parametrizable by two input parameters $\zeta$ and $\zeta'$. The first mimics the human behaviour. What would~one do to escape from a locally optimal solution? She would stop trying to move the endvertices of the edges defining the crossing resolution; she would rather start moving ``irrelevant'' vertices hoping that
%by changing the embedding of the graph or its crossing structure,
by doing so a better solution will be easier to be computed afterwards. Our algorithm is mimicking this idea as follows:
%
\begin{inparaenum}[(i)]
%\item As long as the algorithm detects that within the last $\zeta$ iterations the crossing resolution of the graph has being improved, the vertex-pool contains only critical vertices.
\item if during the last $\zeta$ iterations the crossing resolution has not been~improved, then the vertex-pool becomes \emph{wider} by including all the vertices, and the algorithm is executed with this vertex-pool for $\zeta'$ iterations;
\item afterwards, the vertex-pool switches back to the critical vertices.
\end{inparaenum}
%
%From our extensive practical analysis, we quickly realized that, while
While this approach turned out to be effective for smaller graphs, for graphs with more than 100 vertices, it was not so efficient; in most iterations with the wider vertex-pool, the embedding could not change in a beneficial way for the algorithm~to~proceed.

@@

Instead of computing all crossing points and the corresponding angles for each candidate position of each iteration, we adopted a different approach for determining the set of feasible candidate positions, which turned out to be quite efficient in practice. Recall that we denoted by $v_i$ the vertex chosen at the $i$-th iteration step, and by $\pi_0,\ldots,\pi_{\rho-1}$ the candidate positions to move $v_i$. Let $e_0,\ldots,e_{d_i-1}$ be the edges incident to $v_i$, where $d_i=deg(v_i)$. Next, for each edge $e_k$ with $k=0,\ldots,d_i-1$  we compute the crossings and the corresponding crossing angles of $e_k$ with all other edges in $\Gamma_{i-1}$. Let $\phi_i$ be the minimum crossing angle computed; this is our reference angle. Also, for each candidate position $\pi_j$ with $j=0,\ldots,\rho-1$, and for each edge $e_k$ with $k=0,\ldots,d_i-1$, we compute the crossings and the corresponding crossing angles of $e_k$ with all other edges of the drawing, assuming that $v_i$ is at $\pi_j$. Let $\chi_j$ be the minimum crossing angle computed with this approach, when $v_i$ is at position $\pi_j$. Clearly, $\pi_j$ is feasible only if $\chi_j \geq \phi_i$. Note that the complexity of this approach is $O(deg(v_i)m) = O(nm)$.

@@

In this section, we present the results of our experimental evaluation. For comparison purposes, apart from our algorithm, we also implemented the force-directed algorithms of Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13} and Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}. The implementations\footnote{Our implementation is available on request from the authors.} were in Java using yFiles~\cite{DBLP:books/sp/04/WieseE004}. The experiment was performed on a Linux laptop with four cores at 2.4 GHz and 8 GB RAM.
%
As a test set for our experiment, we used the non-planar Rome graphs~\cite{DBLP:reference/crc/BattistaD13}, which form a collection of around 8.100 benchmark graphs; in Appendix~\ref{app:north}, we also report on the AT\&T graphs.

The experiment was performed as follows. Initially, each Rome graph was laid out using the SmartOrganic layouter of yFiles~\cite{DBLP:books/sp/04/WieseE004}. Starting from this layout, every graph was drawn with
\begin{inparaenum}[(i)]
\item our algorithm,
\item our algorithm restricted~not to violate the aspect ratio of the initial layout, and the force-directed algorithms
\item by Argyriou et al.\ and
\item by Huang et al.
\end{inparaenum}

We compared the quality of the produced drawings based on crossing resolution, total resolution, and angular resolution. Additionally, we consider the aspect ratio, and the number of crossings of the corresponding drawings.

All algorithms of the experiment can easily be adjusted to maximize~only the crossing resolution, or only the angular resolution or both (that is,~the total resolution); see Figs.~\ref{fig:cr-res}, \ref{fig:to-res} and~\ref{fig:an-res}. In our algorithm, this can be achieved by modifying appropriately the content of the vertex-pool (as we saw in Section~\ref{ssec:variants}), while in the algorithms of Argyriou et al.\ and of Huang et al.\ by switching on only the forces that maximize the corresponding properties under measure (note that, each of these two algorithms has a different set of forces to maximize the crossing and the angular resolution, such that together they maximize the total resolution). Additionally, we evaluate the produced drawings regarding the aspect ratio and the number of crossings. The reported results are on average across different drawings with same number of vertices.
